Open-Set Domain Adaptation through Self-Supervision

### 2022-02-10 paper/main.tex
Experiments + Method Rephrase

\subsection{Problem Formulation}
\label{sec:problemformulation}
Our starting point is the source dataset, defined as $\mathcal{D}_s = \{({\bf x}_i^s, y_i^s)\}_{i=1}^{N_S} \sim p_s$, where each element ${\bf x}_i^s$ belonging to any $y_i^s$ is a sample from the source domain $S$.
This dataset has a target counterpart, $\mathcal{D}_t = \{{\bf x}_i^t\}_{i=1}^{N_t} \sim p_t$ which is unlabeled.
In OSDA we have that $p_s \neq p_t$.
The source dataset $\mathcal{D}_s$ is associated with a set of known classes, $\mathcal{C}_s$, which can also be found in the target dataset $\mathcal{D}_t$, but is supposedly smaller.
Hence we have that $|\mathcal{C}_s| < |\mathcal{C}_t|$ and that $\mathcal{C}_s \subset \mathcal{C}_t$.
In a setting of domain adaptation, we further have that $p_t^s \neq p_s$, the target distribution of the known source classes
A metric for measuring how different two domain are is the openness betweeen source and target domain\cite{bendale2015open}, defined as $\displaystyle \mathbb{O} = 1-\frac{\mathcal{C}_s}{\mathcal{C}_t}$.
When $\mathbb{O} > 0$, we're dealing with an OSDA problem.

\subsection{Approach}
\label{sec:apporach}
To tackle the task, we chose to split it in two different steps.
First we have to train the model to separate between the known classes ($\mathcal{C}_s$) and the unknown classes ($\mathcal{C}_{t\setminus s})$ in a reliable enough way.
This is achieved by using a semi-supervised task, by training to model to both recognize a sample class and its correct orientation.
The second step is similar to a classic CSDA problem, where we train the model on a union of both source and target datasets.

@@

Let's denote with $rot({\bf x}, i)$ the rotation of the sample image {\bf x} by $i\times 90$ degrees clockwise.
This is the self-supervised part of our proposed model as rotations $i \in [0, 3]$ can be randomly generated and then predicted.
To avoid situations where the objective orientation of a sample would be a too complicated task, even for a human being, we also feed in input to the model the un-rotated image features.
Alternatively, instead to have the model predict the rotation of a sample for any class, we also try using a different head for each one of the known classes $\mathcal{C}_s$, along with different loss functions.

@@

To separate samples we train the model on an enhanced version of $\mathcal{D}_s$, $\tilde{\mathcal{D}_s} = \{({\bf x}_i^s, {\bf \tilde x}_i^s, z_i^s)\}_{i=1}^{N_s}$ where ${\bf \tilde x}_i^s$ is the rotated version of ${\bf x}_i^s$ and $z_i^s$ is the rotation index associated to image $i$.
The network is composed by an extractor $E$ and two classifiers, $R_1$ and $C_1$ in its standard form.
When using a multi-head rotation classifier, it is composed of $|\mathcal{C}_s|+1$ heads for the rotation task, and an additional one for the classification task.
When using a multi-head predictor, which head to use during separation is chosen by the object classifier $C_1$.
The features of both original and rotated image are used to predict the rotation as ${\bf \tilde z}_s = softmax(R_1([E({\bf x}^s), E({\bf\tilde x}^s)]))$ in the single-head case and using the $j$-th head as ${\bf\tilde{z}}_s = softmax(R_{1,j}[E({\bf x}^s), E({\bf\tilde{x}}^s)])$ in the multi-head case.
Classes are predicted only using un-rotated features as $\tilde{\bf y}^s = softmax(C_1(E({\bf x}_i^s)))$.
The model is training by minimizing an objective function defined as $\mathcal{L}_1 = \mathcal{L}_{C_1} + \mathcal{L}_{R_1}$.
This is the sum of two cross-entropy loss functions as in:

@@

Where $\alpha_1$ is a weight associated to the rotation task.
We also try using an extended rotation objective function $\mathcal{L}_{R_1}^*$ also implementing a center loss function\cite{CenterLoss}:

@@

Here {\bf $v_i$} is the penultimate layer of the rotation classifier, called {\it features}, and $\gamma({\bf z}_i)$ gives the center of the features ${\bf v}_i$ associated to class $i$ and $||\cdot||_2^2$ is the $l$-2 norm and $\lambda_1$ is the weight associated with this extension of the loss function.

When training is completed, we can start separating samples.
To do so, we get the normality score $\mathcal{N}(\cdot)$ which is defined as the maximum prediction of the rotation classifier, $\mathcal{N}({\bf\tilde{x}}_i^s) = \max({\bf\tilde{z}}_i^t)$.
To tell wheter a sample belongs to the known samples of the target domain $\mathcal{D}_t^{knw}$ or the unknown one $\mathcal{D}_t^{unk}$ requires choosing a threshold $\tilde{\mathcal{N}}$.
Then we can simply separate as:

\begin{equation}
  \begin{cases}
    {\bf\tilde x}_i^t \in \mathcal{D}_t^{knw} & \text{ if } \mathcal{N}({\bf\tilde x}_i^s) \geq \tilde{\mathcal{N}} \\
    {\bf\tilde x}_i^t \in \mathcal{D}_t^{unk} & \text{ if } \mathcal{N}({\bf\tilde x}_i^s) < \tilde{\mathcal{N}}
  \end{cases}
  \label{eq:sample_separation}
\end{equation}

When employing a multi-head architecture, we need to choose which one of the $|\mathcal{C}_s|$ heads to use for the classification task.
Head $R_{1,j}$ is used by picking $j$ as $\displaystyle j = \arg\max_j {\bf\tilde{y}}_i^t$.

@@

In this phase, two "new" datasets are used.
The first one is $\mathcal{D}_s^*$, composed as $\mathcal{D}_s \cup \mathcal{D}_t^{unk}$.
Note that since $\mathcal{D}_t$ is an unlabeled dataset but $\mathcal{D}_s^*$ is a labeled one, all samples in $\mathcal{D}_t^{unk}$ are simply labeled as unknowns.
The second one is $\mathcal{D}_t^{knw}$ which is still an unlabeled dataset.
Here we also introduce two new classifiers, $C_2$ and $R_2$.
$C_2$ is the object classifier, but while $C_1$ had a $|\mathcal{C}_s|$-dimensional output, having to accomodate the new "Unknown" class, it has a $|\mathcal{C}_s|+1$-dimensional output.
$R_2$ instead is the rotation classifier, which in this step is always single-headed and has its own weight $\alpha_2$.
These two classifiers try to minimize the objective function described in equations \ref{eq:baseloss_class} and \ref{eq:baseloss_rot}.
One of the main differences between the two steps is that while in the first one we used $C_1$ to regularize the rotation task, here we use $R_2$ to regularize the main classification task.

@@

Evaluating model perfomance requires finding a balance between two values:
{\it OS*}, the share of correctly classified samples;
{\it UNK}, the share of correclty rejected samples.
A model never confident enough to reject a sample as unknown could still achieve high {\it OS*} scores and near-zero {\it UNK}, while a model marking every given sample as unkwnown will achieve perfect {\it UNK} and zero {\it OS*}.
To compare models in a robust manner, we picked an harmonic mean between {\it OS*} and {\it UNK}, defined as $HOS = 2\frac{OS^*\times UNK}{OS^*+UNK}$.
This type of mean is more biased towards the lowest of the two scores, resulting in a more severe evaluation of models.

@@

Our model is tested on the {\it Office-Home} dataset\cite{OfficeHomeDataset}, which features $65$ classes of images over four different domains: Art (A), Clipart (C), Product (P) and Real World (R).
We set the first $45$ classes to be known while the remaining $20$ are unknown.
After each epoch performed during the domain alignment phase, a validation run is performed on the entire original target dataset.
For each experiment, we report both the best achieved HOS and the last reported HOS.
As separation is crucial for the model effectiveness, we also report the computer AUROC score for the first part.

\subsection{Results}
\label{sec:results}
Table \ref{tab:results} contains the results for all 12 (ne abbiamo provati davvero 12???) possible domain shifts, as well as perfomance differences when using multi-head rotation classifiers and center loss.
Each mode has a specific configuration further discussed in \ref{sec:ablation_study}.

>>>>

\subsection{Problem Formulation}
\label{sec:problemformulation}
Our starting point is the source dataset, 
defined as $\mathcal{D}_s = \{({\bf x}_i^s, y_i^s)\}_{i=1}^{N_S} \sim p_s$, 
where each element ${\bf x}_i^s$ belonging to class $y_i^s$ is a sample from the source domain {\it S}.
This dataset has a target counterpart, 
$\mathcal{D}_t = \{{\bf x}_i^t\}_{i=1}^{N_t} \sim p_t$ which is unlabeled instead.
In OSDA we have that $p_s \neq p_t$.
The source dataset $\mathcal{D}_s$ is associated with a set of known classes $\mathcal{C}_s$, 
which can also be found in the target dataset $\mathcal{D}_t$ but is supposedly smaller.
In other words,
$|\mathcal{C}_s| < |\mathcal{C}_t|$ and $\mathcal{C}_s \subset \mathcal{C}_t$.
In domain adaptation, 
we further have that the target distribution of the known source classes differs from the target distribution of known classes,
so $p_t^s \neq p_s$, 
A metric to measure how different two domains are is the openness betweeen source and target domain\cite{bendale2015open}, 
defined as $\displaystyle \mathbb{O} = 1-\frac{\mathcal{C}_s}{\mathcal{C}_t}$.
When $\mathbb{O} > 0$, we're dealing with an OSDA problem.

\subsection{Approach}
\label{sec:apporach}
The task is split in two separate stages.
First, 
we have to train the model to separate between the known classes ($\mathcal{C}_s$) and the unknown classes ($\mathcal{C}_{t\setminus s})$ in a reliable way.
This is achieved via the self-supervised task,
training the model to both recognize the class of a sample and its relative orientation.
Secondly,
we train the model on a union of both source and target datasets,
to reject samples when they are not part of the known classes $\mathcal{C}_s$.

@@

We denote with $rot({\bf x}, k)$ the rotation of the sample image {\bf x} by $k\times 90$ degrees clockwise.
This is the self-supervised part of our proposed model as rotation indices $k \in [0, 3]$ can be randomly generated and then predicted.
Relative rotation,
meaning that both original and rotated features are supplied to the model,
is preferred over just using rotated features as some objects might not have an absolute correct orientation.
Alternatively, 
instead of having the model predict the rotation of a sample for any class using the same classifier, 
we also try using a different head for each one of the known classes $\mathcal{C}_s$.

@@

Separating samples requires iterating over $\tilde{\mathcal{D}_s} = \{({\bf x}_i^s, {\bf \tilde x}_i^s, z_i^s)\}_{i=1}^{N_s}$,
where ${\bf \tilde x}_i^s$ is the $z_i^s\times 90$ degrees rotated version of ${\bf x}_i^s$.
The network is composed by an extractor $E$ and two classifiers, 
$R_1$ and $C_1$.
$C_1$ is the object classifier, 
assigning image labes $\tilde{y}_i^s \in \{0, \dots, |\mathcal{C}_s|\}$,
while $R_1$ is the relative rotation classifier,
assigning rotation indices $\tilde{z}_i^s \in [0, 3]$.
${\bf\tilde{z}}_i^s$ and ${\bf\tilde{y}}_i^s$ are the vector representations of $\tilde{z}_i^s$ and $\tilde{y}_i^s$.
The multi-head rotation classifier uses $|\mathcal{C}_s|$ heads for the rotation task.
In this case, 
which head to use during separation is up to the object classifier $C_1$.
Object predictions are made as ${\bf\tilde{x}}_i^s = softmax(E({\bf x}_i^s))$,
while relative rotation predictions are computed as ${\bf \tilde z}_s = softmax(R_1([E({\bf x}^s), E({\bf\tilde x}^s)]))$.
If a multi-head rotation classifier is used,
the $j = y_i^s$ head is used as ${\bf\tilde{z}}_s = softmax(R_{1,j}[E({\bf x}^s), E({\bf\tilde{x}}^s)])$.
The model is trained to minimizing the objective function $\mathcal{L}_1 = \mathcal{L}_{C_1} + \mathcal{L}_{R_1}$.
This is the sum of two cross-entropy loss functions:

@@

Where $\alpha_1$ is a weight associated to the rotation task.
We also try using an extended rotation loss function $\mathcal{L}_{R_1}^*$ implementing an additional center loss\cite{CenterLoss} term:

@@

Here ${\bf v_i}$ is the penultimate layer of the rotation classifier,
called features, 
$\gamma({\bf z}_i)$ is the center of the features associated to class $i$,
$||\cdot||_2^2$ is the $l$-2 norm and $\lambda_1$ is the weight associated with the center loss term.

When training is completed, 
we can start separating samples.
To do so,
normality scores $\mathcal{N}(\cdot)$ are used,
defined as the maximum prediction of the rotation classifier:
$\mathcal{N}({\bf\tilde{x}}_i) = \max({\bf\tilde{z}}_i)$.
To predict wheter a sample belongs to the known samples of the target domain $\mathcal{D}_t^{knw}$ or the unknown ones $\mathcal{D}_t^{unk}$ requires picking a threshold $\tilde{\mathcal{N}}$.
The separation is then done as:

\begin{equation}
  \begin{cases}
    {\bf x}_i^t \in \mathcal{D}_t^{knw} & \text{ if } \mathcal{N}({\bf\tilde x}_i^t) \geq \tilde{\mathcal{N}} \\
    {\bf x}_i^t \in \mathcal{D}_t^{unk} & \text{ if } \mathcal{N}({\bf\tilde x}_i^t) < \tilde{\mathcal{N}}
  \end{cases}
  \label{eq:sample_separation}
\end{equation}

When using a multi-head rotation classifier, 
it is required to choose among the $|\mathcal{C}_s|$ possible heads to make the prediction.
Head $R_{1,j}$ is used by choosing $j$ as $\displaystyle j = \arg\max_j {\bf\tilde{y}}_i^t$.

@@

In this phase,
we build two "new" datasets.
The first one is $\mathcal{D}_s^*$, composed as $\mathcal{D}_s \cup \mathcal{D}_t^{unk}$.
Since $\mathcal{D}_t$ is an unlabeled dataset but $\mathcal{D}_s^*$ is a labeled one,
samples in $\mathcal{D}_t^{unk}$ are labeled as unknowns.
The second one is $\mathcal{D}_t^{knw}$.
Here we also use two new classifiers,
$C_2$ and $R_2$,
similarly to how they were used in the first phase but with $C_2$ predicting $|\mathcal{C}_s|+1$ classes to accomodate unknown predictions.
In this step $R_2$ is always single-headed and has its own weight $\alpha_2$.
The objective function is the same described in equations \ref{eq:baseloss_class} and \ref{eq:baseloss_rot}.
The main difference here is that $C_2$ is actually trained to recognize unknown classes,
on both domains.

@@

Evaluating model perfomance requires finding a balance between two values:
{\it OS*}, 
the fraction of correctly classified samples,
and {\it UNK}, 
the fraction of correclty rejected samples.
A model not confident enough to reject a sample could still achieve high {\it OS*} values but near-zero {\it UNK},
while a model rejecting every  sample as unkwnown will achieve perfect {\it UNK} and zero {\it OS*}.
To compare models we use the harmonic mean between {\it OS*} and {\it UNK}, 
defined as $HOS = 2\frac{OS^*\times UNK}{OS^*+UNK}$.
This is because this kind of mean tends gives a bigger weight to the smaller values,
resulting in a more severe evaluation of models.

@@

Our model is tested on the {\it Office-Home} dataset\cite{OfficeHomeDataset}, 
which features 65 classes of images over four different domains: Art (A), Clipart (C), Product (P) and Real World (R).
We set the first 45 classes to be known while the remaining 20 are unknown.
For each experiment, 
we report both the best achieved HOS as well as the last reported HOS.
As separation is crucial for the model effectiveness,
 we also report the computer AUROC score for the first part.

\subsection{Results}
\label{sec:results}
Table \ref{tab:results} contains the results for all 12 {\bf (ne abbiamo provati davvero 12???)} available domain shifts,
as well as perfomance differences when using multi-head rotation classifiers and center loss.
Further details are to be found in \ref{sec:ablation_study}.
We can notice that multi-headed models are more capable of correctly discriminating between known and unknown samples,
ultimately resulting in better domain alignment.
A further advantage is given when combined with center loss.
